\chapter{Colimits And Language Extensions}
\label{chap:colimits-language-extension}

This chapter explains the infrastructure developed in:
\begin{itemize}
\item \texttt{Flypitch/Colimit.lean}
\item \texttt{Flypitch/LanguageExtension.lean}
\end{itemize}
The purpose of this layer is to make it possible to compare formulas that live
in different languages, move proofs and models along language maps, and
eventually assemble the infinite Henkin language as a directed colimit.

It builds on the bounded syntax and theory infrastructure from
Chapter~\ref{chap:fol-core} and supplies the technical transport lemmas used
throughout Chapter~\ref{chap:henkinization}.

\section{Directed Diagrams And Their Colimits}

The Henkin construction enlarges the language step by step, so one needs a way
to speak about the limit of a directed family of languages and of the syntax
built over them. The file \texttt{Flypitch/Colimit.lean} develops exactly the
colimit construction needed for this.

\begin{definition}[Directed Type And Directed Diagram]
\label{def:directed-diagram}
A \emph{directed type} is a type equipped with a reflexive, transitive, and
directed relation. A \emph{directed diagram} over such a type assigns an object
to each index and a transition map along each relation, subject to the expected
compatibility law for composition.
\end{definition}

In Lean these are the structures \texttt{directed\_type} and
\texttt{directed\_diagram}. The special case of the natural numbers ordered by
$\le$ is packaged as the definition \texttt{directed\_type\_of\_nat}, together
with a local alias used later in the code.

\begin{definition}[Colimit As Germs]
\label{def:diagram-colimit}
\uses{def:directed-diagram}
Given a directed diagram $F$, the repository first forms the coproduct of all
objects in the diagram. Two elements of this coproduct are declared equivalent
when they eventually map to the same element at some common upper bound. The
colimit is then the quotient by this germ relation.
\end{definition}

The coproduct is \texttt{coproduct\_of\_directed\_diagram}, the relation is
\texttt{germ\_relation}, and the quotient colimit is \texttt{colimit F}. The
theorem \texttt{germ\_equivalence} shows that the germ relation is indeed an
equivalence relation.

\begin{proposition}[Canonical Maps Into The Colimit]
\label{prop:canonical-map-colimit}
\uses{def:diagram-colimit}
Each object of a directed diagram maps canonically into the colimit.
\end{proposition}

\begin{proof}
The map is \texttt{canonical\_map}. The key companion injectivity theorem is:
\begin{itemize}
\item \texttt{canonical\_map\_inj\_of\_transition\_maps\_inj}
\end{itemize}
It says that if all transition maps of the diagram are injective, then every
canonical map into the colimit is injective as well.
\end{proof}

\begin{proposition}[Universal Property]
\label{prop:colimit-universal}
\uses{def:diagram-colimit,prop:canonical-map-colimit}
The quotient colimit carries the expected cocone, and every cocone out of the
diagram induces a unique map from the colimit.
\end{proposition}

\begin{proof}
The canonical cocone is \texttt{cocone\_of\_colimit}, and the induced map is
\texttt{universal\_map}. The theorem
\texttt{universal\_map\_inj\_of\_components\_inj} records a useful injectivity
criterion: if all cocone components are injective, then the universal map is
injective.
\end{proof}

The file ends with a few concrete lemmas for $\omega$-indexed colimits. The
main bookkeeping lemmas are:
\begin{itemize}
\item \texttt{same\_fiber\_as\_push\_to\_sum\_r}
\item \texttt{same\_fiber\_as\_push\_to\_sum\_l}
\end{itemize}
They let later proofs compare two representatives by pushing them to a common
stage.

\section{Language Maps}

The second file introduces explicit maps between first-order languages.

\begin{definition}[Language Homomorphism]
\label{def:language-hom}
\uses{def:fol-language}
A map of languages $L \to L'$ consists of arity-preserving maps on function
symbols and on relation symbols. In Lean this is the structure \texttt{Lhom},
with a dedicated infix notation for language maps in the code.
\end{definition}

The basic operations are all present:
\begin{itemize}
\item identity maps and composition,
\item the inclusions into a sum language,
\item injectivity of a language map, packaged as \texttt{Lhom.is\_injective},
\item decidable range, packaged as \texttt{Lhom.has\_decidable\_range}.
\end{itemize}

The constructions \texttt{on\_term} and \texttt{on\_formula} transport syntax
along a language homomorphism. The file proves that these actions commute with
the structural operations from the first-order core:
\begin{itemize}
\item lifting,
\item substitution,
\item iterated application,
\item composition and identity.
\end{itemize}

This is the syntactic transport layer that later lets a proof written in one
language be reinterpreted inside a larger language.

\section{Tracking Symbols}

Much of the later Henkin argument depends on knowing whether a particular
constant symbol occurs in a term or formula. For that reason the file defines
the set of symbols appearing in a piece of syntax:
\begin{itemize}
\item \texttt{symbols\_in\_term},
\item \texttt{symbols\_in\_formula}.
\end{itemize}

The basic lemmas show that lifting does not change symbol support, while
substitution adds at most the symbols already present in the substituted term.
There are also companion lemmas explaining that if a symbol is not in the range
of a language map, then it does not appear after transporting syntax along that
map.

These support computations are not optional technical clutter. They are the
mechanism that makes ``fresh constant'' arguments precise later on.

\section{Filtering A Language By Symbols}

\begin{definition}[Filtered Language]
\label{def:filtered-language}
\uses{def:language-hom}
Given a predicate $p$ on the symbols of a language $L$, the language
\texttt{filter\_symbols p} keeps exactly the symbols satisfying $p$.
\end{definition}

The canonical map back to the original language is
\texttt{filter\_symbols\_Lhom p}, and it is injective by
\texttt{is\_injective\_filter\_symbols\_Lhom}. The two reconstruction lemmas
\begin{itemize}
\item \texttt{find\_term\_filter\_symbols}
\item \texttt{find\_formula\_filter\_symbols}
\end{itemize}
say that if all symbols appearing in a term or formula satisfy $p$, then that
syntax actually comes from the filtered language.

Conceptually, this lets the development remove a forbidden symbol from the
ambient language and then work entirely inside the smaller language that
contains only the allowed symbols.

\section{Transporting Proofs And Models}

The syntactic action of a language map extends in two directions.

\begin{proposition}[Proof Transport]
\label{prop:on-prf}
\uses{def:language-hom,def:fol-derivability}
A proof in a language $L$ can be pushed forward along a language map
$\varphi : L \to L'$ to obtain a proof in $L'$.
\end{proposition}

\begin{proof}
This is \texttt{on\_prf}, with the sentence-level form \texttt{on\_sprf}. The
inductive proof simply applies \texttt{on\_formula} and \texttt{on\_term} to
every step of a derivation and uses the commutation lemmas for substitution and
quantifiers where needed.
\end{proof}

At the semantic level, a structure on the larger language can be restricted
along a language homomorphism.

\begin{definition}[Reduct]
\label{def:language-reduct}
\uses{def:language-hom,def:fol-structure}
Given $\varphi : L \to L'$ and an $L'$-structure $M$, the \emph{reduct}
\texttt{reduct} of $M$ along $\varphi$ is the $L$-structure obtained by
forgetting the symbols of $L'$ that do not come from $L$.
\end{definition}

The theorem \texttt{reduct\_formula\_iff} states that realizing a transported
formula in $M$ is equivalent to realizing the original formula in the reduct.
At the sentence and theory level this yields
\texttt{reduct\_realize\_sentence} and
\texttt{reduct\_all\_realize\_sentence}.

\section{Reflection Along Injective Maps}

Forward transport is easy. Going backward is subtler: to reflect a proof from a
larger language to a smaller one, one needs to recover preimages of symbols.
That is why the file assumes both injectivity and decidable range.

\begin{definition}[Reflection]
\label{def:language-reflection}
\uses{def:language-hom,def:filtered-language}
For an injective language map $\varphi : L \to L'$ with decidable range, the
definitions \texttt{reflect\_term} and \texttt{reflect\_formula} try to pull
syntax in $L'$ back to syntax in $L$.
\end{definition}

If a symbol of $L'$ lies in the range of $\varphi$, reflection chooses its
preimage. If not, reflection inserts a fresh variable at a designated index.
This replacement-by-variable is the key idea: instead of failing on genuinely
new symbols, reflection records them as open slots.

The long collection of lemmas following these definitions proves that
reflection behaves exactly as needed:
\begin{itemize}
\item it inverts \texttt{on\_term} and \texttt{on\_formula} up to a controlled
lifting operation;
\item it commutes with lifting and substitution;
\item it turns a proof over transported formulas back into a proof in the
original language.
\end{itemize}

The main proof-theoretic output is:
\begin{itemize}
\item \texttt{reflect\_prf}
\item \texttt{reflect\_sprf}
\item \texttt{is\_consistent\_Theory\_induced}
\end{itemize}
Together these show that an injective language map preserves consistency of the
induced theory.

\section{Fresh Constants And Generalization}

The final goal of the chapter is not merely abstract language transport. It is
the fresh-constant generalization package used later in Henkin witness
arguments.

\begin{definition}[Instantiating A Bounded Formula As A Sentence]
\label{def:bounded-subst-sentence}
\uses{def:fol-bounded-syntax}
If $f$ is a bounded formula with one free variable and $t$ is a closed term,
then \texttt{boundedFormulaSubstSentence f t} is the sentence obtained by
substituting $t$ for that variable.
\end{definition}

\begin{theorem}[Generalize A Fresh Constant]
\label{thm:generalize-constant}
\uses{def:filtered-language,def:language-reflection,def:bounded-subst-sentence,prop:on-prf}
Suppose a constant symbol $c$ does not occur in the assumptions $\Gamma$ and
does not occur in a formula $f$. If $\Gamma$ proves the instance obtained by
substituting $c$ into $f$, then $\Gamma$ proves $\forall f$.
\end{theorem}

\begin{proof}
This is the theorem \texttt{generalize\_constant}. The argument removes the
forbidden constant from the language by passing to a filtered language in which
$c$ has been deleted. The assumptions and conclusion are reconstructed inside
that smaller language using the symbol-support lemmas and the reconstruction
theorems:
\begin{itemize}
\item \texttt{find\_term\_filter\_symbols}
\item \texttt{find\_formula\_filter\_symbols}
\end{itemize}

The proof of the instantiated formula is then reflected back into the filtered
language. Because the deleted constant is no longer available there, reflection
replaces it by a fresh variable, exactly producing the universally quantified
formula after one final cancellation lemma for lift-then-substitute.
\end{proof}

\begin{corollary}[Sentence-Level Generalization]
\label{cor:sentence-generalize-constant}
\uses{thm:generalize-constant}
The theorem \texttt{sgeneralize\_constant} packages the same argument for
theories and bounded formulas, which is the form used later in the Henkin
development.
\end{corollary}

\section{Role In The Henkin Construction}

This chapter is the bridge between abstract proof theory and the actual Henkin
machinery.

\begin{itemize}
\item The colimit construction is what later produces the infinite language
\texttt{LInfty}.
\item Language maps let syntax, proofs, and structures move between stages of
the language chain.
\item Reflection and fresh-constant generalization are the technical core of
the witness-adjunction argument.
\end{itemize}

Without this layer, the later statement ``a proof using a fresh constant can be
turned into a universally quantified proof without that constant'' would remain
an informal slogan. Here it becomes a precise and reusable theorem.
