\chapter{Compactness And Completion}
\label{chap:compactness-completion}

This chapter documents the next logic-side step after the first-order core. It
corresponds to the files
\texttt{Flypitch/Compactness.lean} and
\texttt{Flypitch/Completion.lean}. The goal of this layer is to pass from
arbitrary derivations and consistent theories to finite fragments and then to
maximal complete extensions.

It depends on the theory-level notions introduced in
Chapter~\ref{chap:fol-core}, and it provides the compactness and completion
tools used later in Chapter~\ref{chap:henkinization}.

\section{Why Compactness Appears Here}

The later Henkin construction repeatedly enlarges a theory by adding new
sentences. To show that these enlargements preserve consistency, one needs a
way to reduce inconsistency of a large theory to inconsistency of a finite
subtheory. That is exactly the role of compactness in this development.

The compactness argument in the repository is proof-theoretic rather than
semantic. Instead of starting from satisfiability, the code first proves that a
derivation uses only finitely many assumptions, and only then translates this
statement to theories of sentences.

\section{Proof Compactness}

\begin{theorem}[Proof Compactness]
\label{thm:proof-compactness}
If a formula $\psi$ is derivable from a set $T$ of formulas, then there is a
finite subset $\Gamma \subseteq T$ such that $\psi$ is already derivable from
$\Gamma$.
\end{theorem}

\begin{proof}
This is the theorem \texttt{proof\_compactness}. The proof is by induction on a
derivation of $T \vdash' \psi$.

Each inference rule contributes only finitely many assumptions:
\begin{itemize}
\item the axiom rule needs just the singleton containing the cited formula;
\item implication elimination, equality substitution, and similar binary rules
combine the finite supports coming from their premises by taking a finite
union;
\item implication introduction and ex falso remove the discharged assumption by
erasing it from the finite support built for the premise;
\item universal introduction is the only slightly delicate case, because the
premise is proved from the lifted theory \texttt{lift\_formula1 '' $\Gamma$}.
The code rebuilds a finite family of preimages whose lifts are exactly the
finite formulas used in the proof.
\end{itemize}
\end{proof}

This theorem already contains the key finiteness principle needed later: a
contradiction cannot depend on infinitely many assumptions at once.

\section{Theory Compactness}

The theorem above is stated for arbitrary sets of formulas, but the rest of the
development works with theories of sentences. The file therefore repackages the
same statement at the theory level.

\begin{theorem}[Theory Proof Compactness]
\label{thm:theory-proof-compactness}
If a sentence $\psi$ is provable from a theory $T$, then there is a finite set
of sentences $\Gamma \subseteq T$ such that $\psi$ is provable from the finite
theory generated by $\Gamma$.
\end{theorem}

\begin{proof}
The main theorem is \texttt{theory\_proof\_compactness}. The file also records
the equivalent reformulation
\texttt{theory\_proof\_compactness\_iff}. The proof applies
Theorem~\ref{thm:proof-compactness} to the underlying set of formulas
\texttt{Theory.fst T}, then lifts the resulting finite subset back to a finite
set of sentences. The auxiliary definition \texttt{finTheory} packages a
\texttt{Finset} of sentences as a theory.
\end{proof}

This is the form used in consistency arguments: if a theory is inconsistent,
then some finite collection of its sentences is already inconsistent.

\section{Adding New Sentences By Finite Induction}

The file's main compactness-based consistency tool is the theorem
\texttt{is\_consistent\_union}. It expresses a standard but important pattern:

\begin{proposition}[Consistency Of A Controlled Union]
\label{prop:consistent-union}
Let $T_1$ be a consistent theory. Suppose that for every sentence $\psi$ in a
second theory $T_2$, the theory $T_1 \cup \{\neg \psi\}$ proves contradiction.
Then the union $T_1 \cup T_2$ is consistent.
\end{proposition}

\begin{proof}
The Lean theorem is \texttt{is\_consistent\_union}. The proof assumes an
inconsistency in $T_1 \cup T_2$ and applies
Theorem~\ref{thm:theory-proof-compactness} to obtain a finite fragment of that
bad proof.

The finite fragment is then filtered so that only the sentences coming from
$T_2$ remain explicit, while the sentences already in $T_1$ are absorbed into
the base theory. An induction on this finite set shows that one may add its
elements one at a time without losing consistency. At each step the hypothesis
on $T_2$ and the theory-level proof-by-cases lemma
\texttt{sprf\_by\_cases} rule out inconsistency.
\end{proof}

This proposition is the repository's first general mechanism for showing that a
large extension is consistent by checking only one sentence at a time. It will
reappear later in the Henkin chapter.

\section{Chains Of Consistent Extensions}

The completion file studies consistent theories lying over a fixed base theory
$T$. Formally, the type \texttt{TheoryOver T hT} packages a theory extending
$T$ together with a proof of consistency. The ordering relation is inclusion of
underlying theories.

Given a chain of such extensions, the code forms its union and then adjoins the
original base theory. These are the definitions
\texttt{chainTheoryUnion} and \texttt{limitTheory}. The point is that a chain
should have an upper bound, so that Zorn's lemma can later be applied.

\begin{lemma}[Finite Fragments Sit Inside One Chain Element]
\label{lem:finite-subset-limit-theory}
Let $c$ be a nonempty chain of consistent extensions of $T$. Every finite set
of sentences contained in the limit theory of $c$ is already contained in a
single element of the chain.
\end{lemma}

\begin{proof}
This is \texttt{finite\_subset\_limitTheory}. The proof is a finite induction on
the chosen set of sentences. At each step, chain comparability is used to merge
the witness for the old finite set with a witness for the newly added sentence.
\end{proof}

\begin{proposition}[Consistency Of The Limit Theory]
\label{prop:consis-limit}
The union of a chain of consistent extensions of $T$ is again consistent.
\end{proposition}

\begin{proof}
This is the theorem \texttt{consis\_limit}. If the limit theory were
inconsistent, Theorem~\ref{thm:theory-proof-compactness} would produce a finite
inconsistent subtheory. By
Lemma~\ref{lem:finite-subset-limit-theory}, that finite set already lies inside
one chain element, contradicting the consistency stored in that element.
\end{proof}

With this upper-bound construction in hand, the file packages the Zorn
preconditions as \texttt{can\_use\_zorn}.

\section{Maximal Consistent Extensions}

\begin{theorem}[Maximal Consistent Extension]
\label{thm:maximal-extension}
Every consistent theory $T$ has a maximal consistent extension.
\end{theorem}

\begin{proof}
The Lean construction is \texttt{maximal\_extension}. It applies Zorn's lemma
to the partially ordered set of consistent extensions of $T$, ordered by
inclusion. The upper-bound hypothesis required by Zorn is provided by
Proposition~\ref{prop:consis-limit}.
\end{proof}

Maximality is then turned into completeness.

\begin{proposition}[A Maximal Consistent Extension Is Complete]
\label{prop:maximal-is-complete}
If $T_{\max}$ is maximal among the consistent extensions of $T$, then
$T_{\max}$ is complete.
\end{proposition}

\begin{proof}
This is \texttt{complete\_maximal\_extension\_of\_consis}. For a sentence
$\psi$, either $\psi$ is already in $T_{\max}$, or $\neg \psi$ is. If neither
is present, the lemma \texttt{can\_extend} shows that at least one of
$T_{\max} \cup \{\psi\}$ or $T_{\max} \cup \{\neg \psi\}$ remains consistent.
That contradicts maximality.
\end{proof}

\begin{corollary}[Completion Of A Consistent Theory]
\label{cor:completion-of-consistent-theory}
Every consistent theory has a complete consistent extension.
\end{corollary}

\begin{proof}
This is the definition \texttt{completion\_of\_consis}. It simply packages the
maximal extension from Theorem~\ref{thm:maximal-extension} together with
Proposition~\ref{prop:maximal-is-complete}.
\end{proof}

\section{Role In The Overall Development}

This chapter completes the abstract proof-theoretic preparation needed for
Henkinization.

\begin{itemize}
\item Compactness lets the development reduce inconsistency claims to finite
subtheories.
\item Completion turns a merely consistent theory into a complete one.
\item The combination of these two tools is what later makes it possible to add
witnesses systematically while preserving consistency and eventually extract
models from complete Henkin theories.
\end{itemize}
