\chapter{Henkinization}
\label{chap:henkinization}

This chapter documents the implemented Henkin layer in
\texttt{Flypitch/Henkin.lean}. Its role is to enlarge a consistent theory by
adding enough witness constants, then pass to a limit language in which those
witnesses are all available at once.

It combines the completion tools from
Chapter~\ref{chap:compactness-completion} with the language-extension and
colimit machinery from Chapter~\ref{chap:colimits-language-extension}.

The chapter stops exactly where the current Lean 4 development stops. It
documents the henkinization construction itself and the resulting complete
Henkin extension package. It does not pretend that the final completeness
theorem has already been ported.

\section{The One-Step Language Extension}

Given a language $L$, the first operation is to adjoin a new constant symbol
for every bounded formula with one free variable.

\begin{definition}[One-Step Henkin Language]
\label{def:henkin-one-step-language}
\uses{def:fol-bounded-syntax}
The type \texttt{languageFunctions L} adds two kinds of function symbols:
\begin{itemize}
\item the original function symbols of $L$,
\item a nullary witness symbol attached to each bounded formula in one variable.
\end{itemize}
The resulting language is \texttt{languageStep L}.
\end{definition}

The constructor \texttt{wit'} picks out the witness constant associated to a
bounded formula, and \texttt{inclusion} is the canonical map from $L$ into
\texttt{languageStep L}. The theorem \texttt{inclusion\_inj} records that this
language map is injective.

\section{The Omega-Chain Of Languages}

Henkinization does not stop after one step, because new formulas become
available after adding new constants. The file therefore iterates the step
construction.

\begin{definition}[Henkin Language Chain]
\label{def:henkin-language-chain}
\uses{def:henkin-one-step-language,def:language-hom}
The recursively defined languages \texttt{chainObjects L n} are given by:
\begin{itemize}
\item stage $0$ is the original language $L$,
\item stage $n+1$ is \texttt{languageStep} applied to stage $n$.
\end{itemize}
The maps between stages are \texttt{chainMaps}.
\end{definition}

The lemmas \texttt{chainMaps\_comp} and \texttt{chainMaps\_inj} show that these
maps form a directed system of injective language embeddings. This produces the
language-valued directed diagram \texttt{languageChain}.

\section{The Colimit Language \texttt{LInfty}}

The file first redoes the colimit construction at the level of languages,
inside a local namespace specialized to language-valued diagrams. The purpose
is to take the directed colimit of the chain just described.

\begin{definition}[Infinite Henkin Language]
\label{def:linfty-language}
\uses{def:henkin-language-chain,def:diagram-colimit}
The colimit language of the chain \texttt{languageChain} is denoted
\texttt{LInfty L}. The canonical map from stage $m$ into this colimit is
\texttt{canonicalMap m}.
\end{definition}

The theorem \texttt{canonicalMap\_inj} states that each stage embeds
injectively into \texttt{LInfty}. Thus \texttt{LInfty} may be regarded as the
language obtained by adjoining witness constants through all finite stages and
then identifying compatible symbols.

\section{Comparing Stage Syntax With Colimit Syntax}

The central technical task is to show that syntax over \texttt{LInfty} can be
represented by syntax already appearing at some finite stage.

To do this, the file forms directed diagrams not only of languages but also of:
\begin{itemize}
\item terms, via \texttt{termChain},
\item formulas, via \texttt{formulaChain},
\item bounded terms, via \texttt{boundedTermChain},
\item bounded formulas, via \texttt{boundedFormulaChain}.
\end{itemize}

Each of these diagrams carries a cocone into the corresponding syntax over
\texttt{LInfty}. The induced universal maps are:
\begin{itemize}
\item \texttt{termComparison},
\item \texttt{formulaComparison},
\item \texttt{boundedTermComparison},
\item \texttt{boundedFormulaComparison}.
\end{itemize}

\begin{theorem}[Comparison Maps Are Bijective]
\label{thm:henkin-comparison-bijective}
\uses{def:linfty-language,prop:canonical-map-colimit,prop:colimit-universal,def:fol-bounded-syntax}
The four comparison maps above are bijections.
\end{theorem}

\begin{proof}
The Lean theorems are:
\begin{itemize}
\item \texttt{termComparison\_bijective}
\item \texttt{formulaComparison\_bijective}
\item \texttt{boundedTermComparison\_bijective}
\item \texttt{boundedFormulaComparison\_bijective}
\end{itemize}

Injectivity comes from the universal-property injectivity criterion proved in
the colimit chapter, together with injectivity of the canonical language maps.
Surjectivity is more involved: the file recursively reconstructs a finite-stage
representative for every term or formula over \texttt{LInfty}, pushing
subexpressions to a common stage whenever necessary.
\end{proof}

For bounded formulas with one free variable, the resulting bijection is packaged
as the explicit equivalence \texttt{equivBoundedFormulaComparison}. This is the
bridge that later lets the development choose witness constants in
\texttt{LInfty} by descending to a finite stage.

\section{Witness Properties And Enough Constants}

The next layer turns the language construction into a theory construction.

\begin{definition}[Witness Property]
\label{def:henkin-witness-property}
\uses{def:fol-bounded-syntax}
For a bounded formula $f$ with one free variable and a constant $c$, the
sentence \texttt{witProperty f c} says:
\[
\exists x\, f(x) \to f(c).
\]
\end{definition}

Using this sentence, the predicate \texttt{has\_enough\_constants T} expresses
that every bounded one-variable formula has a constant whose witness property is
provable in $T$.

The theorem \texttt{has\_enough\_constants\_of\_subset} shows that this property
is monotone under theory extension. This will matter at the very end, when a
complete extension of the henkinization is produced.

\section{The One-Step Henkin Theory}

\begin{definition}[Henkin Theory Step]
\label{def:henkin-theory-step}
\uses{def:henkin-one-step-language,def:henkin-witness-property}
For a theory $T$ over $L$, write \texttt{henkinTheoryStep T} for the union of:
\begin{itemize}
\item the image of $T$ under the language inclusion into \texttt{languageStep
L},
\item all witness sentences attached to bounded one-variable formulas over $L$.
\end{itemize}
\end{definition}

The file also defines the iterated tower \texttt{henkinTheoryChain T n}. To
state the witness axioms in a proof-friendly form, it introduces:
\begin{itemize}
\item \texttt{henkinWitnessBody},
\item \texttt{henkinWitnessSentence},
\item \texttt{henkinWitnessSentence\_eq\_witProperty}.
\end{itemize}

The point of this reformulation is that the witness sentence is treated as a
specialized instance of the fresh-constant generalization theorem from the
previous chapter.

\section{Consistency Of One Witness Step}

The first genuinely mathematical theorem of the chapter is that one Henkin step
preserves consistency.

\begin{theorem}[Consistency Of One Henkin Step]
\label{thm:henkin-step-consistent}
\uses{def:henkin-theory-step,thm:generalize-constant,thm:theory-proof-compactness}
If $T$ is consistent, then \texttt{henkinTheoryStep T} is consistent.
\end{theorem}

\begin{proof}
This is \texttt{is\_consistent\_henkinTheoryStep}. The proof has three main
ingredients.

\begin{itemize}
\item First, the image of $T$ inside the larger language is consistent by the
consistency-preservation theorem for induced theories.
\item Second, the file isolates freshness lemmas showing that the newly adjoined
witness constant for a formula does not already occur in the relevant theory or
formula body.
\item Third, the theorem \texttt{generalize\_constant} from the language
extension chapter is applied through the helper theorem
\texttt{is\_consistent\_insert\_instance\_of\_exists}.
\end{itemize}

Compactness is then used to reduce a hypothetical inconsistency in the full
Henkin step to a finite collection of witness axioms, which are added one by
one while preserving consistency.
\end{proof}

\section{From The Chain To \texttt{TInfty}}

Once the one-step theorem is in place, the rest of the consistency story is an
induction and colimit argument.

\begin{definition}[Theories Inside \texttt{LInfty}]
\label{def:tinfty-theories}
\uses{def:linfty-language,def:henkin-theory-step}
For each stage $m$, the theory \texttt{iota m} is the image of
\texttt{henkinTheoryChain T m} inside \texttt{LInfty}. The full limit theory is
\texttt{TInfty T}, defined as the union of all these stage images.
\end{definition}

The inclusion lemmas
\texttt{henkinTheoryChainInclusionStep},
\texttt{henkinTheoryChainInclusion}, and
\texttt{iotaInclusionOfLe}
show that these stage theories form an increasing chain.

\begin{theorem}[Consistency Along The Entire Henkin Chain]
\label{thm:henkin-tinfty-consistent}
\uses{thm:henkin-step-consistent,thm:theory-proof-compactness,def:tinfty-theories}
If $T$ is consistent, then all finite Henkin stages are consistent, each
\texttt{iota n} is consistent, the union \texttt{TInfty T} is consistent, and
therefore the henkinization itself is consistent.
\end{theorem}

\begin{proof}
This is spread across the theorems:
\begin{itemize}
\item \texttt{is\_consistent\_henkinTheoryChain}
\item \texttt{is\_consistent\_iota}
\item \texttt{is\_consistent\_TInfty}
\item \texttt{is\_consistent\_henkinization}
\end{itemize}

The proof of \texttt{is\_consistent\_TInfty} again uses compactness. Any finite
inconsistent subtheory of the union already lies inside some single
\texttt{iota n}, contradicting the consistency of that stage.
\end{proof}

\section{Enough Constants In The Limit Language}

The finite-stage witness machinery must still be transferred to
\texttt{LInfty}. This is where the comparison equivalence for bounded formulas
is used.

The definition \texttt{witInfty} chooses, for a bounded one-variable formula
over \texttt{LInfty}, a representative at some finite stage together with the
corresponding witness constant in the next stage. Using this representative, the
theorem \texttt{henkinizationIsHenkin} proves that the consistent theory
\texttt{henkinization} has enough constants.

Conceptually, the argument is:
\begin{itemize}
\item choose a finite-stage representative of the target formula,
\item use the witness axiom already inserted at the next stage,
\item transport that axiom into \texttt{LInfty},
\item identify it with the desired witness property for the original formula.
\end{itemize}

\begin{proposition}[Henkinization Has Enough Constants]
\label{prop:henkinization-has-enough-constants}
\uses{thm:henkin-comparison-bijective,def:henkin-witness-property}
The consistent theory \texttt{henkinization} has enough constants.
\end{proposition}

\begin{proof}
This is the theorem \texttt{henkinizationIsHenkin}. The proof uses the finite-stage
representative chosen by \texttt{witInfty} and transports the corresponding witness
axiom from a finite stage of the Henkin tower into \texttt{LInfty}.
\end{proof}

\section{Complete Henkin Theories Over A Consistent Base}

The file packages the final output as follows.

\begin{definition}[Henkin Language Over A Theory]
\label{def:henkin-language-over}
\uses{def:linfty-language}
The language \texttt{henkinLanguage T} is just \texttt{LInfty} for the base
language of $T$, and \texttt{henkinLanguageOver} is the canonical embedding of
the original language into that limit language.
\end{definition}

The type \texttt{completeHenkinTheoryOver T hT} consists of theory extensions of
$T$ that are both complete and have enough constants.

\begin{theorem}[Complete Henkinization Of A Consistent Theory]
\label{thm:complete-henkinization}
\uses{thm:henkin-tinfty-consistent,cor:completion-of-consistent-theory,prop:henkinization-has-enough-constants,def:henkin-language-over}
Every consistent theory admits a complete Henkin extension over its
henkinization.
\end{theorem}

\begin{proof}
This is \texttt{completeHenkinizationOfConsis}. The proof applies the
completion theorem from the previous chapter to the consistent theory
\texttt{henkinization}. The two supporting inputs are:
\begin{itemize}
\item \texttt{is\_consistent\_henkinization}
\item \texttt{henkinizationIsHenkin}
\end{itemize}
The property of having enough constants is then inherited by the completed
theory by monotonicity under extension.
\end{proof}

\section{Current Boundary}

At this point the Lean 4 project has reconstructed the full henkinization
package needed for the logic branch:
\begin{itemize}
\item the witness-extension language chain,
\item the colimit language \texttt{LInfty},
\item finite-stage representatives for colimit syntax,
\item consistency of the Henkin step and of the entire limit theory,
\item a complete Henkin extension over the consistent henkinization.
\end{itemize}

What remains beyond this chapter is not more basic henkinization machinery. The
next genuine gap is the later completeness-side packaging and then, more
importantly, the forcing and set-theoretic half of the original project.
