\chapter{Status And Next Frontier}
\label{chap:status-next-frontier}

This final chapter summarizes what the current Lean 4 blueprint now documents
and where the project boundary still lies.

\section{What Is Implemented}

The present blueprint covers the currently verified logic-side surface of the
repository:
\begin{itemize}
\item the first-order syntax and proof system,
\item semantic realization and soundness,
\item sentence and theory infrastructure,
\item compactness and maximal consistent extensions,
\item directed colimits and language-extension machinery,
\item the Henkin language chain, \texttt{LInfty}, consistency of
henkinization, and the complete Henkin extension package.
\end{itemize}

Concretely, the verified imported modules are:
\begin{itemize}
\item \texttt{Flypitch/Compat/Core.lean}
\item \texttt{Flypitch/FOL/Syntax.lean}
\item \texttt{Flypitch/FOL/Formula.lean}
\item \texttt{Flypitch/FOL/Proof.lean}
\item \texttt{Flypitch/FOL/Semantics.lean}
\item \texttt{Flypitch/FOL/Theory.lean}
\item \texttt{Flypitch/Compactness.lean}
\item \texttt{Flypitch/Completion.lean}
\item \texttt{Flypitch/Colimit.lean}
\item \texttt{Flypitch/LanguageExtension.lean}
\item \texttt{Flypitch/Henkin.lean}
\item \texttt{Flypitch/Examples/Abel.lean}
\end{itemize}

\section{What This Does Not Yet Mean}

Although the logic branch is now substantial, the full Flypitch port is not yet
complete.

\begin{itemize}
\item The remaining logic-side packaging corresponding to the later
\texttt{completeness} module is not yet documented from Lean 4 code, because
that part is not yet present as verified ported surface.
\item None of the forcing-side modules are present yet.
\item The integration modules culminating in the theorem
\texttt{independence\_of\_CH} are still absent.
\end{itemize}

So the current state should be understood as follows: the repository already
contains most of the internal logic and Henkin machinery needed for the model
theoretic branch, but it does not yet contain the forcing and ZFC constructions
that drive the final independence result.

\section{Current Port Boundary}

According to the project tracker, the next major unported dependency is
\texttt{pSet\_ordinal}. This marks the beginning of the forcing-side branch.

The remaining upstream structure can be grouped into three unfinished layers:
\begin{itemize}
\item the forcing-side foundation beginning with \texttt{pSet\_ordinal},
\item the Boolean-valued model and forcing results,
\item the final integration modules \texttt{bfol}, \texttt{zfc},
\texttt{print\_formula}, and \texttt{summary}.
\end{itemize}

This is why the blueprint currently ends after the Henkin chapter. That is the
last substantial mathematical chapter currently justified by implemented Lean 4
code. The present chapter is therefore a status record rather than the start of
another completed mathematical construction.

\section{How The Blueprint Should Grow Next}

When more modules land, the next documentation passes should follow the same
rule used so far: document only verified content, and organize the exposition by
mathematical construction rather than by raw file order.

In particular, once the forcing side begins to appear, the next natural
blueprint chapters should start from:
\begin{itemize}
\item the \texttt{pSet\_ordinal} groundwork,
\item the topology and regular-open algebra layer,
\item the collapse and Boolean-valued model constructions.
\end{itemize}

\section{Present Reading Guide}

For a reader approaching the current Lean 4 project today, the chapters of this
blueprint should be read as a coherent logic-side story:
\begin{itemize}
\item Chapters~\ref{chap:fol-core} and~\ref{chap:compactness-completion}
develop proof theory, compactness, and completion.
\item Chapter~\ref{chap:colimits-language-extension} explains how syntax and
proofs move across language extensions.
\item Chapter~\ref{chap:henkinization} shows how those tools are assembled into
the henkinization
construction.
\end{itemize}

That story is already mathematically meaningful on its own. It gives a verified
account of the first-order and Henkin-completeness infrastructure on which the
rest of the Flypitch port will eventually build.
